Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-10-24 manu_draft.Rmd
intro replaced with Perry edits

---
title: "Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Perry de Valpine (pdevalpine@berkeley.edu), Rebecca Murphy (rmurphy@chesapeakebay.net), Ian Wren (ianw@sfei.org), Ariella Chelsky (ariellac@sfei.org), Melissa Foley (melissaf@sfei.org), David B. Senn (davids@sfei.org)"
urlcolor: blue
csl: ecology.csl
link-citations: true
---

@@

Many environmental monitoring programs generate fine-scale but potentially irregular time-series data for the purpose of assessing long-term trends for regulatory, management, or scientific purposes.  The mismatch between the scales of monitoring versus analysis questions can lead to statistical challenges.  At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that may not be of direct interest for longer-term trends.  For basic questions, simple smoothing or averaging over time may be appropriate for analyzing longer-term trends. However, simple analyses that aggregate or generalize observations to assess trends can create mismatches between questions of interest and what is returned by the summarized results.  Further, many analysis questions are not easily addressed by simple smoothing methods, such as an assessment of trends that occur at different temporal scales.

In this paper we propose analysis methods that separate the signal-extraction of fine-scale data from the trend analysis at longer time-scales, with full propagation of statistical uncertainties between the two analysis stages.  For the signal-extraction stage, we use smoothing splines, or the larger framework of generalized additive models (GAMs) that encompasses smoothing splines.  While GAMs are a common statistical method and have been used for trend-analysis of environmental monitoring data, alternative and more commonly used approaches can entangle the trend-analysis goal with the signal-extraction goal in the formulation of splines. Analysis questions or management goals that require information about about non-trivial temporal features, such as across-year trends in timing of within-year seasonal events or baseline levels, may not be amenable to a single GAM formulation.  For the trend-analysis stage, error propagation in summary statistics can be addressed using mixed-effects meta-analysis, enabling the application of established tools that incorporate uncertainties in model estimates of summary statistics.  This two-stage approach is useful because across-year time-scales of interest can be imposed or estimated explicitly, which contrasts with implicit absorption of time-scales into spline formulations in other methods.  For example, across-year smoothing methods can be applied to seasonal averages from the signal-extraction stage while accurately propagating uncertainties.  On a practical level, the signal-extraction stage can be viewed as creating a "data product" that can be used in second-stage analyses without the difficulty of embedding every long-term trend question in a GAM modeling framework to model the fine-scale fluctuations.

The smoothing methods under discussion here are distinct from older methods of trend detection, such as seasonal Kendall tests or other non-parametric approaches.  The former has been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years.  The limitations are such that only the magnitude and direction of a change between selected a priori years can be identified, nor are any quantitative descriptions of changes occurring at different temporal scales explicitly provided.  Alternative approaches that decompose a time series into additive or multiplicative components have also been used, such as the seasonal time series composition that isolates a trend, seasonal, and residual component from an observed time series.  Although computationally efficient, such approaches may over-simplify trends into absolute components that do not smoothly vary over time, nor do they allow for explicit hypothesis testing.  More recently, moving window regression approaches have been applied to dynamically estimate parameters that influence temporal changes in water quality time series.  The Weighted Regression on Time, Discharge, and Season (WRTDS) method can describe multi-decadal changes by estimating effects of the predictors that are specific to a given season, year, and flow regime.  The approach provides results that are comparable to non-linear, spline-based methods, although at the expense of computation time and uncertainty estimates that are not easily obtainable.

GAMs have only recently been applied to trend analysis of water quality time-series, particularly in the evaluation of multi-decadal time series from long-term monitoring programs.  The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators.  Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity, including modeling additive annual and seasonal changes and the interaction between the two.  Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).  Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias, addressing serial correlation in high resolution data, or addressing different time lags in describing relationships between response and predictor variables.

Although our approach is not the first to apply GAMs for trend analysis environmental monitoring data, our approach is distinct from previous applications in several ways.  First, some approaches use the GAM smoothing functions to approximate components of a time series to estimate average conditions for a finite period of time.  This can represent a mismatch between processes occurring at different time scales and does not make sense as an estimation method if, for example, a seasonal treatment factor is not an explicit component of the time series.  Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the boundaries of the GAM estimation methods.  There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be confounded.  Perhaps more important, this approach risks obscuring the different time-scales implicit in choices such as an appropriate number of spline knots, leaving the rendering results more difficult to interpret.  Third, over-confidence in conclusions derived from insufficient treatment of model uncertainty, including propogation to secondary analyses, can lead to misinformed decisions.  GAM applications to environmental monitoring data have not fully treated error propagation as a fundamental component of model application to decision-making, particularly when wrong decisions can lead to expensive consequences.

We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Bay, California, USA.  For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data may not adequately describe trends.  Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chloropyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?  We also contrast different GAM formulations that can yield similar signal-extraction results but carry somewhat different interpretations.  The approach we give for using meta-analysis methods as a second stage of analysis following signal extraction with GAMs appears to be new to environmental trend-detection problems.  For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between year.

@@

@@

```{r sumtab, tab.cap = 'Station locations, sample sizes, and median values for salinity, dissolved inorganic nitrogen (DIN, sum of ammonia, nitrate, and nitrite), and chlorophyll.  Rows are arranged from north to south.'}
totab <- rawdat %>% 
  filter(param %in% c('sal', 'din', 'chl')) %>% 
  group_by(station, param) %>% 
  summarise(
    n = n(), 
    medv = median(value, na.rm = T)
  ) %>% 
  ungroup() %>% 
  gather('var', 'val', n, medv) %>% 
  unite('var', param, var) %>% 
  mutate(var = factor(var, levels = c('sal_n', 'sal_medv', 'din_n', 'din_medv', 'chl_n', 'chl_medv'))) %>% 
  left_join(locs, by = 'station') %>% 
  spread(var, val)

tab <- flextable(totab) %>% 
  set_header_labels(
    station = 'Station', 
    lat = 'Latitude', 
    lon = 'Longitude', 
    sal_n = 'n', 
    sal_medv = 'Med.', 
    din_n = 'n', 
    din_medv = 'Med.', 
    chl_n = 'n', 
    chl_medv = 'Med.'
  ) %>% 
  add_header_row(values = c('', '', '', 'Salinity (psu)', '', 'DIN (mg/L)', '', 'Chl-a (ug/L)', '')) %>% 
  merge_at(i = 1, j = 4:5, part = 'header') %>% 
  merge_at(i = 1, j = 6:7, part = 'header') %>% 
  merge_at(i = 1, j = 8:9, part = 'header') %>% 
  colformat_num(j = c(5, 7, 9), digits = 1) %>%
  colformat_num(j = c(2, 3), digits = 3) %>% 
  align(align = 'right', part = 'header') %>% 
  border_remove() %>% 
  hline_top(border = fp_border(color = 'black')) %>% 
  font(fontname = 'Times', part = 'all') %>%
  fontsize(size = 12, part = 'all') %>% 
  width(j = c(2, 3), width = 0.9) 
tab
```

>>>>

---
title: "Multi-scale trend analysis of water quality using error propagation of generalized additive models"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Perry de Valpine (pdevalpine@berkeley.edu), Rebecca Murphy (rmurphy@chesapeakebay.net), Ian Wren (ianw@sfei.org), Ariella Chelsky (ariellac@sfei.org), Melissa Foley (melissaf@sfei.org), David B. Senn (davids@sfei.org)"
urlcolor: blue
csl: ecology.csl
link-citations: true
---

@@

Many environmental monitoring programs generate fine-scale but potentially irregular time-series data to assess long-term trends for regulatory, management, or research purposes. The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18] [Marcus will go over refs]. At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events, management, or seasonal changes).  Such fluctuations may not be of direct interest for longer-term trends or may not be well-suited to multi-scale smoothing methods.  However, aggregate features of seasonal patterns that integrate or summarize short-term fluctuations may be of interest.  In this paper, we describe methods to estimate across-year trends of within-year features of interest such as a seasonal average, seasonal peak, or seasonal timing of events, while accounting for uncertainties across analysis steps.

Previous methods for water quality trend analysis can be generalized into four basic approaches.  Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years [@Hirsch82;@Helsel20].  A literature survey of @Wan17 revealed non-parametric approaches to be the most commonly used methods in long-term water quality trend analysis, yet they have limited scope.  They do not account for changes occurring at different temporal scales, do not adequately evaluate irregularly spaced data [is this true? I would think so.], do not incorporate other explanatory variables, and do not estimate a model that could be useful for other purposes.  Thus, while these non-parametric approaches have some degree of robustness, they apply only to narrow goals.

The seasonal trend decomposition using loess (STL) decomposes a time series into additive components of a long-term trend, a seasonal pattern, and residuals [@Cleveland90;@Cloern10;@Stow15].  While useful and widely applied, this method also has important limitations.  STL decomposition does not incorporate explanatory variables, it is defined more as an algorithm of statistical steps than as a coherent statistical model [e.g., @Wan17], and it does not usually include standard errors to allow hypothesis testing [but see @Hafen10].  Conventional STL approaches may also over-simplify trends into absolute components that do not change over time, e.g., a seasonal estimate that is constant across years.  This limitation presents challenges when addressing questions relevant to long-term water quality data, such as timing of seasonal peaks that can suggest system response to changing  environmental conditions [@Cloern10;@Navarro12].

The more recently developed method of weighted regression on time, discharge, and season (WRTDS) uses a more general local regression scheme than STL [@Hirsch10;@Beck15].  Designed for river data where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allows parameters to vary smoothly in relation to both time and discharge.  This yields parameters that are specific to season, year, and flow regime.  Conceptually, the approach is similar to localized multi-polynomial smoothing methods, although the application was developed specifically for describing long-term water quality trends.  Standard error estimates of predictions are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15]. Although a useful addition to the original method in @Hirsch10, the approach requires extensive resampling as a post-hoc application to a previously fitted model.

The final and most recent approach is to use smoothing splines to separate fluctuations on different time scales and do so within the larger framework of generalized additive models (GAMs) [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19].  These may be seen as generalizing the concepts behind STL and WRTDS. In statistics, the evolution of non-parametric regression methods has largely converged on GAMs rather than more generalized kernel smoothing methods used by both STL and WRTDS.  Kernel smoothing and spline smoothing are closely related, and a key challenge for each is to determine the appropriate degree of smoothing.  For example, WRTDS can potentially give results similar to the spline-based smoothing methods described next, although at higher computational expense and with the limitation that uncertainty estimates are not readily obtainable from the original method [@Beck17]. User input is also required to specify an acceptable degree of smoothing used by the "windows" that define the localized fit of WRTDS at each point in the time series.  These windows are conceptually similar to the kernel (or bandwidth) that used in more conventional smoothing methods. There is no simple rule to guide the choice of defining an appropriate size and a tradeoff between over- and under-smoothing is a hallmark of these approaches.

Compared to kernel smoothing methods, GAMs have various advantages.  They are formulated using "basis functions", and these can be customized for needs such as cyclic splines (e.g., for an annual pattern) and low-dimensional interactions.  They can naturally include both parametric (e.g. linear or quadratic) components and non-parametric (spline) components.  Importantly, multiple approaches to automatically determine the optimal degree of smoothness have been developed, based on likelihood and/or optimizing out-of-sample prediction error.  They have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (GAMMs), and have computationally efficient implementations that can be optimized more quickly than other approaches [@Wood17].  For these reasons, GAMs are widely used for non-parametric regression smoothing in many fields.

GAMs have recently been applied to trend analysis of water quality time-series, particularly from long-term monitoring programs [@Haraguchi15;@Murphy19], but with different formulations and goals than given here. For example, in the US, the Chesapeake Bay Program uses GAMs to decompose time-series into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.  That is related to the methods here, but this paper gives more general methods for analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when they are used as designed [@Wood03;@Wood17] rather than for ad hoc separation of time scales, and prioritizes full incorporation of uncertainty.  Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and different time lags in describing relationships between response and predictor variables [@Lefcheck17;@Testa18].

Our motivating problem has several needs that are not satisfied by previous methods, but can use GAMs as a starting point. Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as Chl-a.  However, the seasonal average within each year must account for different sampling times and intervals, and any trend analysis must incorporate the uncertainties in seasonal averages. STL and/or WRTDS could potentially separate seasonal from long-term trends, but doing so is not necessary to determine seasonal averages.  What is needed is an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, allowing for irregular sampling and the non-independence inherent in time-series.  This can be done with GAMs, but we develop an application that is distinct from previous studies. Even if estimates of seasonal averages and their standard errors are available, none of these methods are designed to understand interannual trends in those averages.  A Kendall test would not incorporate the standard errors or reveal useful long-term patterns beyond a significance test. Similarly, STL and WRTDS are not designed for this goal.

We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Bay, California, USA.  For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data may not adequately describe trends.  Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chlorophyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?  We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] following signal extraction with GAMs that is new to environmental trend-detection problems.  For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between years.

@@

### General approach

The methods proposed here involve three stages.  In the first stage, a GAM is used to estimate a smooth pattern of variation in the raw data along with its uncertainty.  In the second stage, a feature of interest is calculated from the estimated GAM, along with its propagated uncertainty.  In the examples here, the feature is a seasonal average.  Other features could be the timing or magnitude of a seasonal peak, but those are not developed here.  In the third stage, a mixed effects meta-analysis is used to estimate trends and/or test hypotheses about the change in seasonal averages across years.  While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].  They use response data that come with standard errors (uncertainties), which is exactly the need here.  We give two kinds of mixed effects meta-analysis steps: simple comparison of whether seasonal features differ across years, and estimation of short-term linear trends on time-scales chosen by the analyst.  These allow analysis of new kinds of questions for long-term water quality monitoring data.

The three-stage approach is motivated in several ways.  In the first stage, the GAM serves the role of signal-extraction at the scale of the raw data.  We explain the relationships among various GAM specifications and how some previous methods have entangled the trend-analysis goal with the signal-extraction goal in the formulation of splines, making it harder to interpret either one.  The second stage harnesses the capability to propagate uncertainty from estimated GAM parameters to features of interest such as seasonal averages.  Such features are usefully viewed as functions of the fitted GAM rather than parameters to be estimated within the GAM.   On a practical level, the first two stages can be viewed as creating a "data product" that can be used in any subsequent analyses as long as they incorporate the uncertainty associated with each point estimate. This offers an important practical benefit for visualization and management purposes. For the third stage (long-term trend-analysis), we harness existing meta-analysis methods that are well-suited to the task at hand.

@@

```{r sumtab, tab.cap = 'Station locations, sample sizes, and summary values (median, minimum, maximum) for chlorophyll.  Rows are arranged from north to south.'}
totab <- rawdat %>% 
  filter(param %in% c('chl')) %>% 
  group_by(station, param) %>% 
  summarise(
    n = n(), 
    medv = median(value, na.rm = T), 
    minv = min(value, na.rm = T), 
    maxv = max(value, na.rm = T)
  ) %>% 
  ungroup() %>% 
  gather('var', 'val', n, medv, minv, maxv) %>% 
  unite('var', param, var) %>% 
  mutate(var = factor(var, levels = c('chl_n', 'chl_medv', 'chl_minv', 'chl_maxv'))) %>% 
  left_join(locs, by = 'station') %>% 
  spread(var, val)

tab <- flextable(totab) %>% 
  set_header_labels(
    station = 'Station', 
    lat = 'Latitude', 
    lon = 'Longitude', 
    chl_n = 'n', 
    chl_medv = 'Med.', 
    chl_minv = 'Min.', 
    chl_maxv = 'Max.'
  ) %>% 
  # add_header_row(values = c('', '', '', 'Chl-a (ug/L)', '', '', '')) %>% 
  # merge_at(i = 1, j = 4:7, part = 'header') %>% 
  colformat_num(j = c(5, 6, 7), digits = 1) %>%
  colformat_num(j = c(2, 3), digits = 3) %>% 
  align(align = 'right', part = 'header') %>% 
  border_remove() %>% 
  hline_top(border = fp_border(color = 'black')) %>% 
  font(fontname = 'Times', part = 'all') %>%
  fontsize(size = 12, part = 'all') %>% 
  width(j = c(2, 3), width = 0.9) 
tab
```
